Individualized Measures of Partisan Fairness and Utility in Redistricting (??)

### 2021-08-30 paper/paper.Rmd
reorder
typo
add examples and change notation

Let $\Voters$ be the set of voters in the region and time period under consideration, who vote to select a candidate from one of a set of political parties $\Parties$ to represent each of a set of legislative districts, indexed by $\Distrs$.
We take these three sets to be fixed.

Voters vote in elections, each of which produces a specific outcome $e$ from the set $\Elecs$ of all possible election outcomes.
A specific election contains information on how each voter voted.

@@

In an election, voters either vote for a candidate running in their district, or abstain from voting.
Notationally, we have a set of functions $\vote:\Voters\times\Elecs\to \Parties\cup\{\textsc{abstain}\}$.
Votes may be tallied at the district level, either in total or by party: \begin{align*}
    \votes&: \Distrs\times\Elecs \to \N, \qquad 
    \votes(d, e) = \sum_{v\in q^{-1}(d)} \ind_P(\vote(v, e)) \\
    \votes^{(p)}&: \Distrs\times\Elecs\times\Plans \to \N, \qquad 
    \votes^{(p)}(d, e, q) = \sum_{v\in q^{-1}(d)} \ind_{\{p\}}(\vote(v, e)).
\end{align*} The political party of the candidate with the most votes wins each district, as captured by the function $$
    \win: \Distrs\times\Elecs\times\Plans\to P, \qquad 
    \win(d, e, q) = \argmax_{p\in\Parties} \votes^{(p)}(d, e, q).
$$Thus an individual voter $v\in V$ is represented after election $e$ by political party $\win(q(v), e, q)$, and each political party can be said to win a number of seats, formalized as$$
    \seats:\Parties\times\Elecs\times\Plans\to\N, 
    \qquad \seats(p,e,q) = \sum_{d\in\Distrs} \ind_{\{p\}}(\win(d,e,q)).
$$

**[CM: better letter/concept pairings possible?]**

@@

We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
    u^{(G)}:\Voters\times\Elecs\times\Plans\to\R, 
    \qquad u^{(G)}(v,e,q) = g\qty(\frac{\seats(\vote(v, e), e, q)}{|\Distrs|}),
$$for $g:[0,1]\bijto\R$ an increasing function.

@@

*(running examples).*

@@

The specification of the utility of the local election outcome is more straightforward.
We define$$
    u^{(L)}:\Voters\times\Elecs\times\Plans\to\R, 
    \qquad u^{(L)}(v,e,q) = \ind\{\vote(v, e)=\win(q(v), e, q)\},
$$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.

@@

*(running examples).*

@@

The utility measures above are defined for a single election outcome $e\in \Elecs$ and districting plan $q\in\Plans$.
But often we will want to consider utility averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans.

To allow this, we replace the fixed election districting plan with random variables $(E,Q)\sim \pi$, where $\pi$ is a joint probability measure on election outcomes and districting plans.
The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
As we will discuss in Section \ref{sec:estimating}, generally we will assume that $E\indep Q$, though this is not required for our theoretical analysis.
We will also consider a voter $V$ taken uniformly at random from $\Voters$, with $V\indep\pi$.

Then we may define the utilities for a (randomly selected) voter as $$
    U^{(G)} \dfeq u^{(G)}(V, E, Q) \qand
    U^{(L)} \dfeq u^{(L)}(V, E, Q).
$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid V=v$; the electorate's average utility for election outcome $e$ becomes $E[U^{(L)}\mid E=e]$, and so on.

@@

To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
Then we may define the harm to a voter as $h:\Voters\times\Elecs\times\Plans^2\to\R$, \begin{align*}
    h(v,e,q,\tilde q) &= (1 - u^{(L)}(v, e, q))\cdot u^{(L)}(v, e, \tilde q) \\
    &= \ind\{\vote(v, e)\neq\win(q(v), e, q)\}
        \ind\{\vote(v, e)=\win(\tilde q(v), e, \tilde q)\},
\end{align*}that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.

As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
Make the counterfactual plan a random variable $\tilde Q$, with $(E,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
Then the harm to a random voter is $H \dfeq h(V, E, Q, \tilde Q)$.

The expected harm to a specific voter then becomes $$
   \E[H\mid V=v] = \Pr\qty(\vote(v, E)\neq\win(Q(v), E, Q)
    \text{ and }\vote(v, E)=\win(\tilde Q(v), E, \tilde Q)),
$$

@@

*(running examples).*

@@

Continuing with the above notation, the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
    F = \E[H\mid \vote(V, E) = A] - \E[H\mid \vote(V, E) = B];
$$obviously the ordering of $A$ and $B$ is immaterial.
Informally, this quantity is the difference in the probability that a voter of each party is represented by the other party but counterfactually would have been represented by their own party.

Generally, we will want to evaluate the fairness of a specific districting plan $q$ versus a distribution of counterfactuals.
These counterfactuals could arise from a posited probability distribution capturing the factors relevant to a non-partisan redistricting process, or they could be a set of alternative redistricting plans that were considered as part of the legislative process leading to a final enacted plan.
We adopt the notation $F(q)$ for this scenario: $$
    F(q) \dfeq \E[H\mid \vote(V, E) = A, Q=q] - \E[H\mid \vote(V, E) = B, Q=q],
$$ so that we recover $F=\E[F(Q)]$.

@@

*(running examples).*

>>>>

Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate from one of a set of political parties $\Parties$ to represent each of a set of legislative districts, indexed by $\Distrs$.
We take these three sets to be fixed.

Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
In an election, voters either vote for a candidate running in their district, or abstain from voting.
So we write $v:\Voters\to\Parties\cup\{\textsc{abstain}\}$, with $v(i)$ the party that voter $i\in\Voters$ voted for in election $v$.

@@

Votes may be tallied at the district level, either in total or by party: \begin{align*}
    \votes&: \Distrs\times\Elecs \to \N, \qquad 
    \votes(d, v) = \sum_{i\in q^{-1}(d)} \ind_P(v(i)) \\
    \votes^{(p)}&: \Distrs\times\Elecs\times\Plans \to \N, \qquad 
    \votes^{(p)}(d, v, q) = \sum_{i\in q^{-1}(d)} \ind_{\{p\}}(v(i)).
\end{align*} The political party of the candidate with the most votes wins each district, as captured by the function $$
    \win: \Distrs\times\Elecs\times\Plans\to P, \qquad 
    \win(d, v, q) = \argmax_{p\in\Parties} \votes^{(p)}(d, v, q).
$$Thus an individual voter $i\in\Voters$ is represented after election $v$ by political party $\win(q(i), v, q)$, and each political party can be said to win a number of seats, formalized as$$
    \seats:\Parties\times\Elecs\times\Plans\to\N, 
    \qquad \seats(p,v,q) = \sum_{d\in\Distrs} \ind_{\{p\}}(\win(d,v,q)).
$$

@@

We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
    u^{(G)}:\Voters\times\Elecs\times\Plans\to\R, 
    \qquad u^{(G)}(i,v,q) = g\qty(\frac{\seats(v(i), v, q)}{|\Distrs|}),
$$for $g:[0,1]\bijto\R$ an increasing function.

@@

@@

The specification of the utility of the local election outcome is more straightforward.
We define$$
    u^{(L)}:\Voters\times\Elecs\times\Plans\to\R, 
    \qquad u^{(L)}(i,v,q) = \ind\{v(i)=\win(q(i), e, q)\},
$$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.

@@

## Examples

To illustrate these two utility measures, Figure \ref{fig:mini-util-opt} shows three plans for our running examples which maximize both global (with $g=\log$) and local utility.[^2]
In Mininois and Minissouri, the overall division of seats is as close to proportional as possible.
In Minichusetts, however, it is impossible to give any seats to Republicans (as long as districts are constructed from precincts); all plans have the same utility.
In none of the three states is it possible to create a perfect bipartisan gerrymander, but the plans for Mininois and Minissouri get as close as possible by packing Democratic voters into cities (and Republican voters into rural areas, where feasible).

[^2]: These examples present a slight complication---what to do when a party wins no seats?
    The easiest solution, perhaps, is to define the utility to that party's voters to be $-1$; this will make the utility well-defined while still encouraging proportionality.

```{r mini-util-opt, fig.cap="Utility-maximizing redistricting plans for the three running examples, with each district shaded by its partisan lean."}
include_graphics(here("paper/figures/minis_util_opt.pdf"))
```

@@

The utility measures above are defined for a single election outcome $v\in \Elecs$ and districting plan $q\in\Plans$.
But often we will want to consider utility averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans.

To allow this, we replace the fixed election districting plan with random variables $(V,Q)\sim \pi$, where $\pi$ is a joint probability measure on election outcomes and districting plans.
The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V\indep Q$, though this is not required for our theoretical analysis.
We will also consider a voter $I$ taken uniformly at random from $\Voters$, with $I\indep\pi$.

Then we may define the utilities for a (randomly selected) voter as $$
    U^{(G)} \dfeq u^{(G)}(I, V, Q) \qand
    U^{(L)} \dfeq u^{(L)}(I, V, Q).
$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $e$ becomes $\E[U^{(L)}\mid V=v]$, and so on.

@@

To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
Then we may define the harm to a voter as $h:\Voters\times\Elecs\times\Plans^2\to\R$, \begin{align*}
    h(i,v,q,\tilde q) &= (1 - u^{(L)}(i, v, q))\cdot u^{(L)}(i, v, \tilde q) \\
    &= \ind\{v(i)\neq\win(q(i), e, q)\}
        \ind\{v(i)=\win(\tilde q(i), e, \tilde q)\},
\end{align*}that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.

As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
Make the counterfactual plan a random variable $\tilde Q$, with $(V,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
Then the harm to a random voter is $H \dfeq h(I, V, Q, \tilde Q)$.

The expected harm to a specific voter then becomes $$
   \E[H\mid I=i] = \Pr\qty(V(i)\neq\win(Q(i), V, Q)
    \text{ and }V(i)=\win(\tilde Q(i), V, \tilde Q)),
$$

@@

```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm)."}
include_graphics(here("paper/figures/mini_harm.pdf"))
```

Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure \ref{fig:mini-util-opt}.
As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

@@

Continuing with the above notation, the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
    F = \E[H\mid V(I) = A] - \E[H\mid V(I) = B];
$$obviously the ordering of $A$ and $B$ is immaterial.
Informally, this quantity is the difference in the probability that a voter of each party is represented by the other party but counterfactually would have been represented by their own party.

Generally, we will want to evaluate the fairness of a specific districting plan $q$ versus a distribution of counterfactuals.
These counterfactuals could arise from a posited probability distribution capturing the factors relevant to a non-partisan redistricting process, or they could be a set of alternative redistricting plans that were considered as part of the legislative process leading to a final enacted plan.
We adopt the notation $F(q)$ for this scenario: $$
    F(q) \dfeq \E[H\mid V(I) = A, Q=q] - \E[H\mid V(I) = B, Q=q],
$$ so that we recover $F=\E[F(Q)]$.

@@

In the districting plan examples of Figure \ref{fig:mini-harm}, the fairness of the Democratic gerrymander of Mininois is $-0.494$, the fairness of the Democratic gerrymander of Minissouri is $-0.415$, and the fairness of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
In other words, the average Republican voter in Mininois is 49.4% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
